Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2012-10-22 documents/mixtures/mixture_models.tex
figure reference
notes to self, text, and MoG acronym

% to-do
% -----
% - HOGG: can we drop the regularization in the badness and still succeeed?  If so, let's do that (and drop it in the text).
% - DSTN: make figures showing use of the models in a real situation; see last figure
% - HOGG: does the Lupton paper give the profiles, or should we cite the code directly?
% - HOGG: need Bundy reference
% - HOGG: make what I say about M=6 lux and M=8 lux true -- ie, edit the Tractor code!
% - HOGG: send to Lupton and Bundy for comments
% - DSTN: can we be more specific about how Bosch is using them? I would like to cite users.
% - DSTN: is there anyone we need to ack on your side, or granting agencies?
% - HOGG: should we make Machine Readable Tables for PASP?

@@

\begin{abstract}
Exponential, de~Vaucouleurs, and S\'ersic profiles are simple and
successful models for fitting two-dimensional images of galaxies.  One
numerical issue encountered in this kind of fitting is the pixel
rendering and convolution (or correlation) of the models with the
telescope point-spread function (PSF); these operations are slow, and
easy to get slightly wrong at small radii.  Here we exploit the
realization that these models can be approximated to arbitrary
accuracy with a mixture (linear superposition) of two-dimensional
Gaussians.  Mixtures of Gaussians are fast to render and fast to
affine-transform.  Most importantly, if you have a mixture-of-Gaussian
PSF model (which we advocate), the PSF-convolved, affine-transformed
galaxy models are themselves mixtures of Gaussians and therefore very
fast to compute, integrate, and render precisely.  We present worked
examples that can be directly used in image fitting; we are using them
ourselves.  The mixture profiles we provide can be swapped in to
replace the standard models in any image-fitting code; they ought to
make that code faster at very little cost in accuracy.
\end{abstract}

@@

To elaborate, a mixture of multivariate Gaussians---a linear
superposition---can be used to represent any reasonable distribution
in any number of dimensions to any reasonable precision.  Convolution
(or correlation) by any other distribution that has also been
represented by a mixture of multivariate Gaussians creates a new
mixture of Gaussians with simply adjusted amplitudes, means, and
variance tensors.  The ubiquity of convolution operations in astronomy
suggests the widespread adoption of mixture-of-Gaussian modeling.  We
have pioneered this in the area of distribution modeling in large
numbers of dimensions (\citealt{xd, xdqso, xdqsoz}), where convolution
occurs because the true or noise-free distribution is convolved with
the noise before being observed.  Here we are going to capitalize on
the convolution properties of mixtures of Gaussians in modeling galaxy
morphologies in imaging data, where the true or
high-angular-resolution intensity field is convolved with the
point-spread function (PSF) before being observed.  This is always
easier with Gaussians than cuspy profiles, but is \emph{particularly}
useful in data sets in which the PSF itself has also been modeled as a
mixture of Gaussians, which is not uncommon (for example, the
\project{Sloan Digital Sky Survey} imaging pipelines make approximate
mixture-of-Gaussian PSF models for every imaging field;
\citealt{lupton}).

We are not the first in this space:  Deconvolution and modeling of
galaxy images with mixtures of Gaussians has been done very
successfully before (for example, \citealt{bendinelli, emsellem,
  bendinelli2, cappellari}).  However, the idea behind those projects
was to use the mixture of Gaussians to provide a very free form for
the isophotes or morphologies of resolved galaxies or other complex
scenes.  Here our goals are very limited: We want to improve the
performance of standard galaxy image model fitting by expressing the
standard galaxy models---the exponential and de~Vaucouleurs
profiles---as rigid mixtures of Gaussians.

@@

The performance advantages we will obtain from using
mixture-of-Gaussian approximations are not simply that convolution
itself is trivial.  The standard unconvolved galaxy models, especially
the de~Vaucouleurs and S\'ersic models (\citealt{dev}, \citealt{ser}),
are very ill-behaved near the galaxy center.  Rendering these profiles
precisely near the center can be very challenging numerically.  In
addition, most of the rendering time is spent at very small radii,
where PSF convolution is going to erase all structure anyway.  That
is, a mixture-of-Gaussians description of the de~Vaucouleurs profile
saves time both in rendering \emph{and} in PSF convolution; it
produces profiles that are approximate but very high performance when
all real uses of the profiles are PSF-convolved, as they usually are.

Along those lines, it is important in any image-modeling situation to
think of the PSF as the \emph{pixel-convolved} point-spread function.
Under this choice, synthesis of a pixelized image involves only
convolution with the PSF and sampling at the pixel centers.  That is,
image synthesis (returned to at the end of this \documentname)
consists of PSF convolution (an arithmetic operation on the two
mixtures of Gaussians) followed by sampling (evaluation of the
resulting mixture at the pixel centers).

@@

Because we are thinking about two-dimensional imaging, we use here
two-dimensional Gaussian or Normal distributions, which look like
\begin{eqnarray}\displaystyle
\normal(\pos\given\mean,\var) &\equiv& \frac{1}{2\pi}\,\det(\var)^{-1/2}\,\exp(-\frac{1}{2}\,\transpose{[\pos-\mean]}\cdot\inverse{\var}\cdot[\pos-\mean])
\quad ,
\end{eqnarray}
where $\pos$ and $\mean$ are two-dimensional vectors (usually in the
focal plane or on the sky or something like that), and $\var$ is a
symmetric $2\times 2$ variance tensor or matrix, and implicitly the
vectors are column vectors.  A \emph{mixture of Gaussians} is a linear
superposition of Gaussians.  Any positive two-dimensional function
with finite support and finite total integral---including as a special
case any two-dimensional probability distribution function---can be
represented as a mixture of Gaussians to arbitrary accuracy; that is
\begin{eqnarray}
p(\pos) &\approx& \sum_{k=1}^K a_k\,\normal(\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K a_k
\quad ,
\end{eqnarray}
where $p(\pos)$ is any probability distribution of a two-dimensional
quantity $\pos$, the $\approx$ symbol implies approximation, $K$ is
the number of Gaussians used in the mixture, and the $K$ Gaussians
have amplitudes $a_k$, means $\mean_k$, and variance tensors $\var_k$.
The sum-to-one condition ensures that the probability distribution
approximation is properly normalized.

We wish to make an approximation to the two-dimensional circular
exponential (exp) profile $Q^{\exp}(\cdot)$
\begin{eqnarray}\displaystyle
Q^{\exp}(\spos) &\equiv& \exp(-\alpha^{\exp}\,[|\spos| - 1])
\\
\alpha^{\exp} &\equiv& 1.67834699
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and
$\alpha^{\exp}$ is a dimensionless inverse length set to ensure that
the profile has unit half-light radius.  The position $\spos$ is
dimensionless because it parameterizes the unit-size dimensionless
function.  We seek the best (where ``best'' will be defined below)
$M^{\exp}$-Gaussian mixture (where $M^{\exp}$ is an integer)
approximation
\begin{eqnarray}\displaystyle
Q^{\exp}(\spos) &\approx& \sum_{m=1}^{M^{\exp}} a^{\exp}_m\,\normal(\spos\given\zero,\var^{\exp}_m)
\\
\var^{\exp}_m &\equiv& v^{\exp}_m\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\var^{\exp}_m$ in the mixture can be represented as a scalar
$v^{\exp}_m$ multiplied by the identity matrix $\identity$
because we are requiring this dimensionless function to be precisely
circular (so every component is itself circular and concentric).
Similarly for the de~Vaucouleurs (dev) profile
\begin{eqnarray}\displaystyle
Q^{\dev}(\spos) &\equiv& \exp(-\alpha^{\dev}\,[|\spos|^{1/4} - 1])
\\
\alpha^{\dev} &\equiv& 7.66924944
\\
Q^{\dev}(\spos) &\approx& \sum_{m=1}^{M^{\dev}} a^{\dev}_m\,\normal(\spos\given\zero,\var^{\dev}_m)
\\
\var^{\dev}_m &\equiv& v^{\dev}_m\,\identity
\quad .
\end{eqnarray}
The half-light inverse-radius parameters $\alpha^{\exp}$ and
$\alpha^{\dev}$ are from \citet{ciotti}.  The challenge we
meet below is to determine the parameters
\begin{eqnarray}
\{a^{\exp}_m,v^{\exp}_m\}_{m=1}^{M^{\exp}}~,~\{a^{\dev}_m,v^{\dev}_m\}_{m=1}^{M^{\dev}}
\end{eqnarray}
to best approximate the traditional galaxy profile functions, under
some sensible definiton of the word ``best'', as a function of the
model complexity parameters (numbers of components) $(M^{\exp},
M^{\dev})$.

@@

The profiles above are normalized to have unit intensity
(approximately) at their half-light radii.  In many cases, the
investigator wants profiles that are normalized to have unit total
flux (intensity integrated over solid angle).  Although there is an
analytic result for the dev profile, numerical integration of the
concentrated profiles dev and luv to determine total fluxes can be
challenging.  This is not true for the mixture-of-Gaussian
approximations: Each Gaussian is normalized, so the sum of the
amplitudes $\sum_m a^{\luv}_m$ (for the luv profile, say) gives the
total flux for the mixture-of-Gaussians approximation to that profile.

We seek the \emph{best} mixture-of-Gaussian approximations.  This
necessitates definition of the word ``best''.  If we think of the
profiles as being two-dimensional probability distribution functions
(for, say, the arrivals of photons), then one natural choice is the
K-L divergence or similar cross-entropy or information-theoretic
measure.  However, in typical astronomical imaging, the galaxy is
superimposed on a substantial, flat sky level, and the noise in the
data is close to Gaussian.  This suggests more chi-squared-like
objectives.  We adopt the latter, in part because they are most
appropriate for our specific proposed application (modeling \sdss-like
astronomical imaging), but experiments we have performed suggest that
information-theoretic objectives also lead to good results.

In detail, the chi-squared objective we minimize---the
\emph{badness}---is a squared residual between the exact profile
function $Q(\spos)$ and its mixture-of-Gaussian approximation.  It is
designed to be equivalent to a chi-squared statistic in a
homoskedastic two-dimensional image of the profile taken with
extremely high angular resolution (pixels of size 0.001 the half-light
radius) and vanishing point-spread function.  Quantitatively the
badness is defined to be the mean squared residual in the $Q(\spos)$
functions, which are normalized to have unit intensity at the
half-light radius, averaged over a two-dimensional circular region in
the $\spos$ plane centered on the (circularly symmetric) profile and
extending out to radius $\xi_{\max}$.  We use $\xi_{\max}=8$ for all
profiles except the $\lux$ profile, for which we use $\xi_{\max}=4$.
In practice, the badness is computed in a one-dimensional numerical
integral but the integral is weighted in radius (weight increasing
linearly with radius) to make it equivalent to the two-dimensional
chi-squared.  We also add to the badness a very tiny coefficient (on
the order of $10^{-3}$ of the best-fit badness) times the sum of the
variances $v_m$ for regularization.  In practice, this term doesn't
have much effect and could be dropped.

@@

In our work on \project{The Tractor}, we use the $M^{\lux}=6$ lux and
the $M^{\luv}=8$ luv profiles.  Our advice to users would be to do the
same.  We use the lux and luv over the exp and dev partly because of
their better behaviors numerically, and partly because they and we are
both part of the \project{SDSS} tradition.  These---$M^{\lux}=6$ and
$M^{\luv}=8$---are good compromises between mixture complexity ($M$)
and quality of fit (badness).  Also, even the best-fitting late-type
and early-type galaxies deviate from exponential and de~Vaucouleurs
fits by more than do these high-quality mixture-of-Gaussian
approxmations; no precision is lost.

The value of these mixture-of-Gaussian galaxy components comes when
they are to be convolved with a PSF (usually in fact a pixel-convolved
PSF) that is itself represented as a mixture of Gaussians.  In this
scenario, the PSF $\psi(\Delta\pos)$---which is thought of as a
function of focal-plane displacement $\Delta\pos$ away from, say, a
true stellar position---is represented as a $K$-Gaussian mixture
\begin{eqnarray}\displaystyle
\psi(\Delta\pos) &=& \sum_{k=1}^K p_k\,\normal(\Delta\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K p_k
\quad ,
\end{eqnarray}
where the means $\mean_k$ are \emph{not} required to vanish because
the PSF can have arbitrarily non-trivial structure (think speckles and
the like) and the variances $\var_k$ will not in general be
proportional to the identity or even diagonal because the PSF will not
in general be round.  An example that illustrates the use of this PSF
is the following: A star of flux $S_s$ at focal-plane position
$\pos_s$ will lead to an image (PSF-convolved intensity map) of the
form
\begin{eqnarray}\displaystyle
I(\pos\given\star,S_s,\pos_s) &=& \sum_{k=1}^K S_s\,p_k\,\normal(\pos\given\pos_s+\mean_k,\var_k)
\quad .
\end{eqnarray}
That is, when the PSF is represented as a mixture of Gaussians, any
image of a star---or indeed any image of any set of stars---is also
represented as a mixture of Gaussians.

Applying this PSF to an exp or dev galaxy is slightly more
complicated, because the galaxy has not just a flux $S_g$ and a
central position $\pos_g$; it also has a shape.  Because we are only
considering these simple galaxies, we are only permitting ellipsoidal
shapes, which can be represented by a semi-major axis $a$, a
semi-minor axis $b$, and a position angle $\phi$, or equivalently by
eigenvalues $a, b$ and eigenvectors $\uv_1, \uv_2$, or equivalently by
an affine transformation $\affine_g$ that takes a circle to the
relevant ellipse (and is a therefore a general representation of an
ellipse; it is also the matrix square root of the symmetric variance
tensor describing the ellipse).  The galaxy is distorted by this
affine transformation \emph{prior} to PSF convolution, so the
focal-plane image (PSF-convolved intensity field) for a general (say)
$\exp$ galaxy is given by
\begin{eqnarray}\displaystyle
I(\pos\given\exp,S_g,\pos_g,\affine_g) &=& \sum_{k=1}^K \sum_{m=1}^{M^{\exp}} S_g\,a^{\exp}_m\,p_k\,\normal(\pos\given\pos_g+\mean_k,\var_{gm}+\var_k)
\\
\var_{gm} &\equiv& \affine_g\cdot\var^{\exp}_m\cdot\transpose{\affine_g}
\\
\affine_g &=& \left[a\,\uv_1 , b\,\uv_2 \right]
\quad ,
\end{eqnarray}
where $a$ and $b$ are the major and minor axis lengths of the galaxy
ellipse (in appropriate units) and $\uv_1$ and $\uv_2$ are the
eigenvectors on the sky pointing in the major-axis and minor-axis
directions respectively.  Implicitly all vectors are two-dimensional
column vectors, and $\affine_g$ is a $2\times 2$ affine transformation
matrix that contains the ``shape'' (position angle, major-axis, and
ellipticity) information about the galaxy.  The $\dev$, $\luv$, and
$\lux$ cases are all similar.  Note the important and key result of
this \documentname, to wit, that a mixture-of-Gaussian galaxy model
(with $M$ components) convolved with a mixture-of-Gaussian PSF model
(with $K$ components) yields a mixture-of-Gaussian image (with
$[M\,K]$ components).

DSTN... pegagogical figure demonstrating these operations and their results...

@@

One amusing aspect of mixture-of-Gaussian profiles has to do with
projection from three to two dimensions.  The projection of a
three-dimensional Gaussian is a two-dimensional Gaussian; the
two-dimensional, rigid, circular approximations we have made for the
ser profiles can be deprojected to rigid, spherical approximations to
the three-dimensional profiles trivially.  The two-dimensional models
we have started with are not accurate models of galaxies in
detail---no galaxy follows exactly any ser profile---so deprojection
of our approximations are are not that interesting in themselves.
However, the general program of fitting two-dimensional sources with
mixtures of Gaussians may have strong implications in the future for
three-dimensional modeling and deprojection.

>>>>

% to-do
% -----
% - DSTN: we need *some* estimate of the speed-up this delivers to The Tractor in the abstract and in the text -- of course this is bullshit because we get to decide how precisely we were doing things prior to using the mixtures -- but we need it since it is our main point.
% - HOGG: can we drop the regularization in the badness and still succeeed?  If so, let's do that (and drop it in the text).
% - DSTN: make figures showing use of the models in a real situation; see last figure
% - HOGG: does the Lupton paper give the profiles, or should we cite the code directly?
% - HOGG: need Bundy reference
% - HOGG: make what I say about M=6 lux and M=8 lux true -- ie, edit the Tractor code!
% - HOGG: send to Lupton and Bundy for comments
% - DSTN: can we be more specific about how Bosch is using them? I would like to cite users.
% - DSTN: is there anyone we need to ack on your side, or granting agencies?
% - HOGG: should we make Machine Readable Tables for PASP?

@@

\begin{abstract}
Exponential, de~Vaucouleurs, and S\'ersic profiles are simple and
successful models for fitting two-dimensional images of galaxies.  One
numerical issue encountered in this kind of fitting is the pixel
rendering and convolution (or correlation) of the models with the
telescope point-spread function (PSF); these operations are slow, and
easy to get slightly wrong at small radii.  Here we exploit the
realization that these models can be approximated to arbitrary
accuracy with a mixture (linear superposition) of two-dimensional
Gaussians (MoGs).  MoGs are fast to render and fast to
affine-transform.  Most importantly, if you have a MoG PSF model
(which we advocate), the PSF-convolved, affine-transformed galaxy
models are themselves MoGs and therefore very fast to compute,
integrate, and render precisely.  We present worked examples that can
be directly used in image fitting; we are using them ourselves.  The
MoG profiles we provide can be swapped in to replace the standard
models in any image-fitting code; they sped up model fitting in our
projects by XXX; they ought to make any code faster at very little
cost in accuracy.
\end{abstract}

@@

To elaborate, a mixture of multivariate Gaussians---a linear
superposition---can be used to represent any reasonable distribution
in any number of dimensions to any reasonable precision.  Convolution
(or correlation) by any other distribution that has also been
represented by a mixture of multivariate Gaussians creates a new
mixture of Gaussians with simply adjusted amplitudes, means, and
variance tensors.  The ubiquity of convolution operations in astronomy
suggests the widespread adoption of mixture-of-Gaussian (MoG)
modeling.  We have pioneered this in the area of distribution modeling
in large numbers of dimensions (\citealt{xd, xdqso, xdqsoz}), where
convolution occurs because the true or noise-free distribution is
convolved with the noise before being observed.  Here we are going to
capitalize on the convolution properties of MoGs in modeling galaxy
morphologies in imaging data, where the true or
high-angular-resolution intensity field is convolved with the
point-spread function (PSF) before being observed.  This is always
easier with Gaussians than cuspy profiles, but is \emph{particularly}
useful in data sets in which the PSF itself has also been modeled as a
MoG, which is not uncommon (for example, the \project{Sloan Digital
  Sky Survey} imaging pipelines make approximate MoG PSF models for
every imaging field; \citealt{lupton}).

We are not the first in this space: Deconvolution and modeling of
galaxy images with MoGs has been done very successfully before (for
example, \citealt{bendinelli, emsellem, bendinelli2, cappellari}; they
use the name ``multi-Gaussian expansion'' or MGE).  However, the idea
behind those projects was to use the MoGs to provide a very free form
for the isophotes or morphologies of resolved galaxies or other
complex scenes.  Here our goals are very limited: We want to improve
the performance of standard galaxy image model fitting by expressing
the standard galaxy models---the exponential and de~Vaucouleurs
profiles---as rigid MoGs.

@@

The performance advantages we will obtain from using MoG
approximations are not simply that convolution itself is trivial.  The
standard unconvolved galaxy models, especially the de~Vaucouleurs and
S\'ersic models (\citealt{dev}, \citealt{ser}), are very ill-behaved
near the galaxy center.  Rendering these profiles precisely near the
center can be very challenging numerically.  In addition, most of the
rendering time is spent at very small radii, where PSF convolution is
going to erase all structure anyway.  That is, a MoG description of
the de~Vaucouleurs profile saves time both in rendering \emph{and} in
PSF convolution; it produces profiles that are approximate but very
high performance when all real uses of the profiles are PSF-convolved,
as they usually are.

Along those lines, it is important in any image-modeling situation to
think of the PSF as the \emph{pixel-convolved} point-spread function.
Under this choice, synthesis of a pixelized image involves only
convolution with the PSF and sampling at the pixel centers.  That is,
image synthesis (returned to at the end of this \documentname)
consists of PSF convolution (an arithmetic operation on the two MoGs)
followed by sampling (evaluation of the resulting MoG at the pixel
centers).

@@

Because we are thinking about two-dimensional imaging, we use here
two-dimensional Gaussian or Normal distributions, which look like
\begin{eqnarray}\displaystyle
\normal(\pos\given\mean,\var) &\equiv& \frac{1}{2\pi}\,\det(\var)^{-1/2}\,\exp(-\frac{1}{2}\,\transpose{[\pos-\mean]}\cdot\inverse{\var}\cdot[\pos-\mean])
\quad ,
\end{eqnarray}
where $\pos$ and $\mean$ are two-dimensional vectors (usually in the
focal plane or on the sky or something like that), and $\var$ is a
symmetric $2\times 2$ variance tensor or matrix, and implicitly the
vectors are column vectors.  A mixture of Gaussians is a linear
superposition of Gaussians.  Any positive two-dimensional function
with finite support and finite total integral---including as a special
case any two-dimensional probability distribution function---can be
represented as a MoG to arbitrary accuracy; that is
\begin{eqnarray}
p(\pos) &\approx& \sum_{k=1}^K a_k\,\normal(\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K a_k
\quad ,
\end{eqnarray}
where $p(\pos)$ is any probability distribution of a two-dimensional
quantity $\pos$, the $\approx$ symbol implies approximation, $K$ is
the number of Gaussians used in the MoG, and the $K$ Gaussians
have amplitudes $a_k$, means $\mean_k$, and variance tensors $\var_k$.
The sum-to-one condition ensures that the probability distribution
approximation is properly normalized.

We wish to make an approximation to the two-dimensional circular
exponential (exp) profile $Q^{\exp}(\cdot)$
\begin{eqnarray}\displaystyle
Q^{\exp}(\spos) &\equiv& \exp(-\alpha^{\exp}\,[|\spos| - 1])
\\
\alpha^{\exp} &\equiv& 1.67834699
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and
$\alpha^{\exp}$ is a dimensionless inverse length set to ensure that
the profile has unit half-light radius.  The position $\spos$ is
dimensionless because it parameterizes the unit-size dimensionless
function.  We seek the best (where ``best'' will be defined below)
$M^{\exp}$-Gaussian MoG (where $M^{\exp}$ is an integer) approximation
\begin{eqnarray}\displaystyle
Q^{\exp}(\spos) &\approx& \sum_{m=1}^{M^{\exp}} a^{\exp}_m\,\normal(\spos\given\zero,\var^{\exp}_m)
\\
\var^{\exp}_m &\equiv& v^{\exp}_m\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\var^{\exp}_m$ in the MoG can be represented as a scalar
$v^{\exp}_m$ multiplied by the identity matrix $\identity$
because we are requiring this dimensionless function to be precisely
circular (so every component is itself circular and concentric).
Similarly for the de~Vaucouleurs (dev) profile
\begin{eqnarray}\displaystyle
Q^{\dev}(\spos) &\equiv& \exp(-\alpha^{\dev}\,[|\spos|^{1/4} - 1])
\\
\alpha^{\dev} &\equiv& 7.66924944
\\
Q^{\dev}(\spos) &\approx& \sum_{m=1}^{M^{\dev}} a^{\dev}_m\,\normal(\spos\given\zero,\var^{\dev}_m)
\\
\var^{\dev}_m &\equiv& v^{\dev}_m\,\identity
\quad .
\end{eqnarray}
The half-light inverse-radius parameters $\alpha^{\exp}$ and
$\alpha^{\dev}$ are from \citet{ciotti}.  The challenge we
meet below is to determine the parameters
\begin{eqnarray}
\{a^{\exp}_m,v^{\exp}_m\}_{m=1}^{M^{\exp}}~,~\{a^{\dev}_m,v^{\dev}_m\}_{m=1}^{M^{\dev}}
\end{eqnarray}
to best approximate the traditional galaxy profile functions, under
some sensible definiton of the word ``best'', as a function of the
model complexity parameters (numbers of components) $(M^{\exp},
M^{\dev})$.

@@

The profiles above are normalized to have unit intensity
(approximately) at their half-light radii.  In many cases, the
investigator wants profiles that are normalized to have unit total
flux (intensity integrated over solid angle).  Although there is an
analytic result for the dev profile, numerical integration of the
concentrated profiles dev and luv to determine total fluxes can be
challenging.  This is not true for the MoG approximations: Each
Gaussian is normalized, so the sum of the amplitudes $\sum_m
a^{\luv}_m$ (for the luv profile, say) gives the total flux for the
MoG approximation to that profile.

We seek the \emph{best} MoG approximations.  This necessitates
definition of the word ``best''.  If we think of the profiles as being
two-dimensional probability distribution functions (for, say, the
arrivals of photons), then one natural choice is the K-L divergence or
similar cross-entropy or information-theoretic measure.  However, in
typical astronomical imaging, the galaxy is superimposed on a
substantial, flat sky level, and the noise in the data is close to
Gaussian.  This suggests more chi-squared-like objectives.  We adopt
the latter, in part because they are most appropriate for our specific
proposed application (modeling \sdss-like astronomical imaging), but
experiments we have performed suggest that information-theoretic
objectives also lead to good results.

In detail, the chi-squared objective we minimize---the
\emph{badness}---is a squared residual between the exact profile
function $Q(\spos)$ and its MoG approximation.  It is designed to be
equivalent to a chi-squared statistic in a homoskedastic
two-dimensional image of the profile taken with extremely high angular
resolution (pixels of size 0.001 the half-light radius) and vanishing
point-spread function.  Quantitatively the badness is defined to be
the mean squared residual in the $Q(\spos)$ functions, which are
normalized to have unit intensity at the half-light radius, averaged
over a two-dimensional circular region in the $\spos$ plane centered
on the (circularly symmetric) profile and extending out to radius
$\xi_{\max}$.  We use $\xi_{\max}=8$ for all profiles except the
$\lux$ profile, for which we use $\xi_{\max}=4$.  In practice, the
badness is computed in a one-dimensional numerical integral but the
integral is weighted in radius (weight increasing linearly with
radius) to make it equivalent to the two-dimensional chi-squared.  We
also add to the badness a very tiny coefficient (on the order of
$10^{-3}$ of the best-fit badness) times the sum of the variances
$v_m$ for regularization.  In practice, this term doesn't have much
effect and could be dropped.

@@

In our work on \project{The Tractor}, we use the $M^{\lux}=6$ lux and
the $M^{\luv}=8$ luv profiles.  Our advice to users would be to do the
same.  We use the lux and luv over the exp and dev partly because of
their better behaviors numerically, and partly because they and we are
both part of the \project{SDSS} tradition.  These---$M^{\lux}=6$ and
$M^{\luv}=8$---are good compromises between mixture complexity ($M$)
and quality of fit (badness).  Also, even the best-fitting late-type
and early-type galaxies deviate from exponential and de~Vaucouleurs
fits by more than do these high-quality MoG approxmations; no
precision is lost.

The value of these MoG galaxy components comes when they are to be
convolved with a PSF (usually in fact a pixel-convolved PSF) that is
itself also represented as a MoG.  In this scenario, the PSF
$\psi(\Delta\pos)$---which is thought of as a function of focal-plane
displacement $\Delta\pos$ away from, say, a true stellar position---is
represented as a $K$-Gaussian MoG
\begin{eqnarray}\displaystyle
\psi(\Delta\pos) &=& \sum_{k=1}^K p_k\,\normal(\Delta\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K p_k
\quad ,
\end{eqnarray}
where the means $\mean_k$ are \emph{not} required to vanish because
the PSF can have arbitrarily non-trivial structure (think speckles and
the like) and the variances $\var_k$ will not in general be
proportional to the identity or even diagonal because the PSF will not
in general be round.  An example that illustrates the use of this PSF
is the following: A star of flux $S_s$ at focal-plane position
$\pos_s$ will lead to an image (PSF-convolved intensity map) of the
form
\begin{eqnarray}\displaystyle
I(\pos\given\star,S_s,\pos_s) &=& \sum_{k=1}^K S_s\,p_k\,\normal(\pos\given\pos_s+\mean_k,\var_k)
\quad .
\end{eqnarray}
That is, when the PSF is represented as a MoG, any image of a
star---or indeed any image of any set of stars---is also represented
as a MoG.

Applying this PSF to an exp or dev galaxy is slightly more
complicated, because the galaxy has not just a flux $S_g$ and a
central position $\pos_g$; it also has a shape.  Because we are only
considering these simple galaxies, we are only permitting ellipsoidal
shapes, which can be represented by a semi-major axis $a$, a
semi-minor axis $b$, and a position angle $\phi$, or equivalently by
eigenvalues $a, b$ and eigenvectors $\uv_1, \uv_2$, or equivalently by
an affine transformation $\affine_g$ that takes a circle to the
relevant ellipse (and is a therefore a general representation of an
ellipse; it is also the matrix square root of the symmetric variance
tensor describing the ellipse).  The galaxy is distorted by this
affine transformation \emph{prior} to PSF convolution, so the
focal-plane image (PSF-convolved intensity field) for a general (say)
$\exp$ galaxy is given by
\begin{eqnarray}\displaystyle
I(\pos\given\exp,S_g,\pos_g,\affine_g) &=& \sum_{k=1}^K \sum_{m=1}^{M^{\exp}} S_g\,a^{\exp}_m\,p_k\,\normal(\pos\given\pos_g+\mean_k,\var_{gm}+\var_k)
\\
\var_{gm} &\equiv& \affine_g\cdot\var^{\exp}_m\cdot\transpose{\affine_g}
\\
\affine_g &=& \left[a\,\uv_1 , b\,\uv_2 \right]
\quad ,
\end{eqnarray}
where $a$ and $b$ are the major and minor axis lengths of the galaxy
ellipse (in appropriate units) and $\uv_1$ and $\uv_2$ are the
eigenvectors on the sky pointing in the major-axis and minor-axis
directions respectively.  Implicitly all vectors are two-dimensional
column vectors, and $\affine_g$ is a $2\times 2$ affine transformation
matrix that contains the ``shape'' (position angle, major-axis, and
ellipticity) information about the galaxy.  The $\dev$, $\luv$, and
$\lux$ cases are all similar.  Note the important and key result of
this \documentname, to wit, that a MoG galaxy model (with $M$
components) convolved with a MoG PSF model (with $K$ components)
yields a MoG model image (with $[M\,K]$ components).

In the above we said ``pixel-convolved PSF''.  In every context, when
modeling images, it is valuable to use the pixel-convolved PSF.  With
this definition of the PSF, the pixelized image is the PSF-convolved
true model evaluated at the pixel centers.  This operation is fast.
Other definitions for the PSF (the non-pixel-convolved, for example)
require that the user to \emph{two} convolutions, the first withe the
PSF and the second with the square (or worse) pixel.  Our advice: Only
fit and use pixel-convolved PSFs.

DSTN... In \figurename~\ref{fig:example} we show...

If your PSF is not in MoG form, it is still the case that convolution
of a MoG approximation to a dev (say) profile will in general be
easier than convolution of the original dev profile.  The reason is
that convolution of a Gaussian with any PSF is fast (indeed most
image-processing languages have such functions built in); the
PSF-convolved profile becomes in this case just a mixture of
Gaussian-convolved PSFs.

The speed-ups that can be obtained by using MoG approximations can be
very large... DSTN: How much did we benefit when we switched to MoGs
in \project{The Tractor}?...

@@

One amusing aspect of MoG profiles has to do with projection from
three to two dimensions.  The projection of a three-dimensional
Gaussian is a two-dimensional Gaussian; the two-dimensional, rigid,
circular approximations we have made for the ser profiles can be
deprojected to rigid, spherical approximations to the
three-dimensional profiles trivially.  The two-dimensional models we
have started with are not accurate models of galaxies in detail---no
galaxy follows exactly any ser profile---so deprojection of our
approximations are are not that interesting in themselves.  However,
the general program of fitting two-dimensional sources with MoGs may
have strong implications in the future for three-dimensional modeling
and deprojection.
